A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-09 paper.tex
Pass through ancestry resolution section

\section*{Ancestry resolution}\label{Ancestry_resolution}
% Need to introduce Hudson's algorithm as concerned with simulating the process of
% the coalescent with recombination as efficiently as possible, and properly merge in
% with this section.

The edge annotations in Figs~\ref{fig-arg-data-structure} and \ref{fig-arg-in-pedigree}
describe the regions of genome that have been passed from a parent to its immediate children,
hence edges either span the entire genome, or entire region to the left (or right)
of a breakpoint
(e.g. Fig~\ref{fig-arg-data-structure}C). In this form of edge annotation, which we call
``immediate edge annotation'',  % NB - perhaps we can think of a better term than this
it is inefficient to extract the local trees from the ARG structure.

There is, however, an alternative edge annotation scheme which allows efficient
storage and extraction of the local trees. It is based on the fact that the ARG
structure is inherently retrospective, defined as the ancestry of a set of \emph{sample}
nodes (such as \textsf{a}, \textsf{b}, and \textsf{c} in the examples above). We define a
``sample-resolved edge annotation'' to be a representation where edge spans
represent only those parts of the genome that have been inherited by the samples.
This also allows the removal of nodes which are not genetically ancestral to the samples.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][fig.~1]{wiuf1999recombination} example graph as a gARG, with contrasting
edge annotation schemes.
(A) Immediate edge annotation, reflecting ``local'' transmission of all
genetic material between parent and child. All edges span the full genome from position 0 to 7,
apart from those linking a recombinant and its two parents (bold).
(B) Sample-resolved edge annotation: the Hudson-like simplification algorithm has been used to
shorten edge spans to reflect only the upwards transmission of genetic
material from the samples \textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within
each node is shaded as in figure \ref{fig-arg-in-pedigree}a to show
coalescing regions of genome back to the grand MRCA genome \textsf{w}.
Line widths reflect the genomic span of each edge, which
helps visualise the relative importance of different transmission pathways.
}
\end{figure}

Fig~\ref{fig-ancestry-resolution} illustrates the two different edge annotation
schemes. To obtain the sample-resolved annotation in
Fig~\ref{fig-ancestry-resolution}B we must traverse backwards in time from
sample nodes \textsf{a}, \textsf{b}, and \textsf{c}, using the edge intervals
to record regions of sample genomes that have passed through each route. When
edges join into a parent node, the parent is associated with the union of the
child intervals. When a child node is connected to two parents, one parent is
allocated the edge intervals of the child, truncated to include only genomic
material to the left of the breakpoint. Similarly the other parent is allocated
the child's intervals truncated such that they contain only regions to the
right of the breakpoint (a natural extension exists for multiple chromosomal
breakpoints).

For example, in Fig~\ref{fig-ancestry-resolution}A the edge between child node \textsf{g} and
parent node \textsf{r} spans the entire genome from 0..7, reflecting the material
transmitted immediately between parent and child.
However, in Fig~\ref{fig-ancestry-resolution}B the edge has been reduced in extent so that
it now spans 0..2, the region that eventually ends up in
sample node \textsf{c}. This sample-resolved edge annotation scheme cannot be
derived locally: it must be done by traversing backwards in time from the sample nodes
up through the ARG.

% TECHNICAL DESCRIPTION
% Given a gARG $E$ and the set of samples $S$ we can \emph{resolve} its
% ancestry by noting that the resolved interval $Ir$ on an edge $(c,p,I)$
% is given by [Rough notes]
% \begin{itemize}
% \item The ``ancestry interval'' $A_u$ of a node $u$ is the union of
% all $R_{u, v}$ for all the children $v$ of $u$. (this isn't quite right)
% \item The resolved interval is then the intersection of $A_u$ and the
% original inheritance interval $I_v$ on the edge.
% \item It would be good of if differentiate between the inheritance intervals
% from the gARG and the ancestry intervals of the resolved gARG.
% \end{itemize}
% This process is illustrated in Fig~\ref{fig-ancestry-resolution}. It is intimately linked to
%Hudson's algorithm which operates by
% tracking segments of ancestral material carried by ancestors as
% we go backwards in time.

% jk-note unclear here about the best choice of notation. We could have
% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
% stick with the original tuple notation for now.
%
% Each lineage consists of a set of
% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
% such that for each $1 \leq j \leq \left| l \right|$,
% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
% $\mathbf{a}_j$ is an integer
% tracking the number of samples the lineage is ancestral to over the interval.
% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
% covered by the lineage, respectively.

When traversing upwards through the ARG like this, it is also helpful to record
the number of sample nodes to which a given section of genome is ancestral.
This is indicated in Fig~\ref{fig-ancestry-resolution}B by successively darker shading
of the regions within each node. When a region has all samples as descendants, it is
shaded black, and we can say that that region of the genome has fully coalesced.

Formally, we can define a sample-resolved gARG as a tuple $(S, E)$
where $S \subset \mathbb{N}$ is a set of sample nodes and
$E$ is a set of edges. Each element of $E$
is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$
over which genome $c$ inherits from $p$, and there is a path from
$c$ to some non-empty subset of $S$ through the local trees.

The ARG traversal algorithm which implements this edge trimming and removal process
is a form of ``simplification'' (cite), and can be thought of as running Hudson's algorithm
on a fixed tree topology (see appendix XXX).

Several features of Fig 3B are worthy of note. First, since node \textsf{j} is not on
the ancestral path from the samples it (and its associated edges) can be removed.
For larger ARGs, the absence of these nodes makes sample-resolved edge annotation
substantially more efficient than immediate edge annotation.
Second, by the time we reach
nodes \textsf{t} and \textsf{v}, all regions of the genome have fully coalesced, and so
\textsf{w}, the ``grand MRCA'' can be omitted from the graph, at the cost of having
multiple root nodes in different regions of the genome.
Third, there exist nodes such as
\textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
the breakpoint, although they retain the information that a breakpoint occurred between
positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).

% Should probably define how we build the local trees somewhere.
TODO: explain why this encoding allows efficient extraction of local trees, as implemented in the tskit library.

% We should talk about ARG and HUD here. Fig~\ref{fig-ancestry-resolution}A shows
% ARG (the Griffiths graph). However, we have a small issue that subfig B
% doesn't strictly show HUD, because it doesn't terminate as soon as all samples have coalesced.
% We could potentially colour the top (non-HUD) section of the plot in a different colour?

TODO: give some feel for how much of a reduction the sample-resolved encoding can give in a large ARG, and
maybe mention ``big'' vs ``little'' args as may be described in Jere's appendix

% Potentially useful text from a previous iteration:
% Wiuf and Hein's classical pair of papers
% clarified the relationship between Hudson's algorithm and the ARG.
% ARG simulation is inherently less efficient than Hudson's
% algorithm because we must track all lineages, even those that correspond
% to sections that have coalesced, or we won't have a complete ARG.

@@

The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root (see the
``\nameref{Ancestry_resolution}'' section).
Such unary nodes in a local tree are not usually considered important,
because they provide very little information and are not
detectable by information from any given individual tree.
The nodes in a tree usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be of equivalent status to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative and (for most computational purposes) inefficient nodes
from the local trees and the corresponding annotated graph.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The best
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
and  \textsf{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by a single edge. This definition includes the case
of a single node that has one inbound and one outbound edge, such as
nodes \textsf{f} and \textsf{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

>>>>

\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
The edges in a gARG represent a path of genetic inheritance from
parent to child (and ultimately from cell to cell) through some
number of generations. The edges are ``annotated'' (see section XXX)
with the coordinates over which inheritance occurs. There are
two ways in which we can interpret these annotations, and the
distinction is subtle and important.
The first interpretation is to define the annotations as
the genetic intervals inherited by the descendant node from the
ancestor node. The second is to view the edge annotations as representing
the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample that is propagated through
the graph from descendant to ancestor.
These alternatives are illustrated in Fig~\ref{fig-ancestry-resolution}.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][Fig.~1]{wiuf1999recombination} example graph as a gARG,
with contrasting
edge annotation schemes. (A) Immediate edge annotation, reflecting ``local''
transmission of all genetic material between parent and child. All edges span
the full genome from position 0 to 7, apart from those linking a recombinant
and its two parents (bold). (B) Sample-resolved edge annotation: the
Hudson-like simplification algorithm has been used to shorten edge spans to
reflect only the upwards transmission of genetic material from the samples
\textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within each
node is shaded as in Fig.~\ref{fig-arg-in-pedigree}A to show coalescing
regions of genome back to the grand MRCA \textsf{w}. Line widths reflect
the genomic span of each edge, which helps visualise the relative importance of
different transmission pathways.
}
\end{figure}

ARGs are inherently a retrospective structure defined in terms
of a set of sample nodes (usually the leaves), and,
as we discuss in section XXX, storing sample-resolved ancestry
intervals (Fig~\ref{fig-ancestry-resolution}B) is generally much more
useful than storing the local parent-child intervals of inheritance
(Fig~\ref{fig-ancestry-resolution}A).
Indeed, storing local parent-child
inheritance intervals for edges is really only of interest because of
its direct correspondence with the classical approach of associating
breakpoints with recombination nodes in an eARG. The eARG encoding
is simply a more limited form, in which the only types of direct
inheritance we can have are the entire genome, or the intervals
at either side of a recombination breakpoint.

Fig~\ref{fig-ancestry-resolution}A is an instructive example
from the literature \citep{wiuf1999recombination} and illustrates
how a classical eARG can be described as a gARG with local edge annotations
(directly corresponding to the eARG description), and how we can
then derive the ancestry resolved edge annotations.
It is important to note, however, that the example is a sample from
the so-called ``big'' ARG \emph{process} (see Appendix XXX)
and has several features that we are unlikely to infer in practice.
For example, by definition the samples contain no genetic information about
node \textsf{j} and so it is unlike to be inferred.
Likewise, is is unlikely we would have information to infer the ``grand MRCA''
\textsf{w} as all regions of the genome have already reached a
common ancestor separately
(in nodes \textsf{o}, \textsf{u}  and \textsf{v}).

% Third, there exist nodes such as
% \textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
% but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
% within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
% node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
% the breakpoint, although they retain the information that a breakpoint occurred between
% positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).

Given the local inheritance edge annotations
(Fig~\ref{fig-ancestry-resolution}A),
 we obtain the sample-resolved edge annotation
(Fig~\ref{fig-ancestry-resolution}B) by traversing the graph backwards in time
from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
Proceeding in topological order (such that an ancestor node is visited
after all of its descendants) the ancestral material for a node is the
union of the ancestral intervals on its inbound edges. We then annotate
the outbound edges with the intersection of their local inheritance and
just-computed ancestral intervals for the node.
For example, node \textsf{d} in Fig~\ref{fig-ancestry-resolution}
carries ancestral material on the interval $(0, 4]$, since it
is the left-hand parent of the recombinant \textsf{b}. The edge
joining \textsf{d} to its ancestor \textsf{h} then carries ancestral
material on $(0, 4]$ since although \textsf{d} inherits the entire interval
$(0, 7]$ from \textsf{h}, only $(0, 4]$ is ancestral to the sample.
This process is directly analogous to Hudson's simulation algorithm
(Appendix XXX).

These alternatives of recording the direct inheritance of genetic material from
parent to child vs the propagation of genomic intervals ancestral to the sample
are analogous to running forwards-time vs backwards-time simulations.
Indeed, the process of taking a graph with edges annotated with direct parent-child
inheritance (Fig~\ref{fig-ancestry-resolution}A) and producing the equivalent
``ancestry resolved'' graph in which only those genomic intervals that are
ancestral to the sample are retained (Fig~\ref{fig-ancestry-resolution}B)
is the same key algorithm underlying recent advances in forwards-time simulation.
\cite{kelleher2018efficient} developed a method in which forwards-time
simulations are performed by recording all parent-child genetic relationships
(i.e.~annotating the edges in the simulated pedigree with the genomic
intervals inherited by a child from a parent) and periodically ``simplifying''
this structure to retain only ancestral relationships that are relevant
to the ancestry of the current population. In many situations, recording
the full genetic ancestry of the population in this way is orders of magnitude faster
than the alternative of tracking
sequences~\citep{kelleher2018efficient,haller2018tree}.

@@

The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a
(backwards-time) Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root (see the
``\nameref{Ancestry_resolution}'' section).
Such unary nodes in a local tree are not usually considered important;
% because they provide little information and are not
% detectable by information from any given individual tree.
nodes usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be of equivalent status to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative and (for most computational purposes) inefficient nodes
from the local trees and the corresponding annotated graph.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The best
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
and  \textsf{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by one edge. This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \textsf{f} and \textsf{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.
